---
title: "Integrating analysis of scRNA-seq and bulk RNA-seq data"
author: "Lin Yang"
date: "2026-03-12"
format: html
editor: visual
---

This test aims to infer the cell type proportion of bulk RNA-seq samples based on real scRNA-seq data from the same tissue which provides us more accurate information of the cell composition than traditional way by using estimate or CIBERSORT.

### datasets used in this test

#### scRNA-seq

-   source: GSE149614
-   original samples: single-cell transcriptomes for 10 HCC patients from four relevant sites: primary tumor, portal vein tumor thrombus (PVTT), metastatic lymph node and non-tumor liver
-   filtered samples: cells from primary tumor

#### bulk RNA-seq

-   source: TCGA-LIHC cohort
-   samples: 377 LIHC samples and 59 control samples

#### Those steps are included in this run:

-   scRNA-seq data quality control and preprocessing
-   scRNA-seq data normalization and cell clustering
-   scRNA-seq data cell annotation
-   bulk RNA-seq data clean (gene symbol conversion, etc)
-   cell proportion comparison between tumor and control samples
-   bulk RNA-seq DEG analysis
-   GSEA analysis

```{r message=FALSE, warning=FALSE}
library(Seurat)
library(Matrix)
library(ggplot2)
library(dplyr)
library(tibble)
library(SingleCellExperiment)
library(scuttle)      
library(patchwork)
library(data.table)
library(clusterProfiler)
library(org.Hs.eg.db)
library(stringr)
library(tidyverse)
library(ggpubr)
library(RColorBrewer)
library(patchwork)
library(psych)
library(corrplot)
library(DESeq2)
library(ggrepel)
library(fgsea)  

```

#### scRNA-seq data processing

```{r}
setwd("/Users/linyang/Documents/sc-bulk_RNA-seq")
raw_count_mat <-
  read.table("GSE149614_HCC.scRNAseq.S71915.count.txt.gz",
          row.names = 1,stringsAsFactors = F)
raw_meta_mat <- read.table("GSE149614_HCC.metadata.updated.txt.gz",
           stringsAsFactors = F,
           header=T)
row.names(raw_meta_mat) <- raw_meta_mat$Cell

raw_count_mat <- as.matrix(raw_count_mat)

counts <- Matrix(raw_count_mat, sparse = TRUE)

seurat_obj <- CreateSeuratObject(
  counts = counts,
  meta.data = raw_meta_mat,
  project = "scRNA_project",
  min.cells = 3,
  min.features = 200
)
sc_data <- subset(seurat_obj,site=="Tumor") ## only to keep cells from primary tumor
sc_data[["percent.mt"]] <- PercentageFeatureSet(sc_data, pattern = "^MT-") # filter on mitochondrial RNA expression
sc_data[["percent.ribo"]] <- PercentageFeatureSet(sc_data, pattern = "^RP[SL]") # filter on ribosomal RNA expression
sc_data
```

```{r}
# check on the detected mRNA counts distribution
sc_data@meta.data %>%
  ggplot(aes(x=nFeature_RNA))+
  geom_histogram(binwidth=20)
```

```{r}
# check on the detected read counts distribution
sc_data@meta.data %>%
  ggplot(aes(x=nCount_RNA))+
  geom_histogram(binwidth=20)
```

```{r}
sc_data[["log10GenesPerUMI"]] <- log10(sc_data$nFeature_RNA) / log10(sc_data$nCount_RNA)
qc_features <- c("nFeature_RNA", "nCount_RNA", "percent.mt", 
                  "percent.ribo", "log10GenesPerUMI")

pre_filter_violin <- VlnPlot(
  sc_data, 
  features = qc_features,
  pt.size  = 0,      # hide points for large datasets
  ncol     = 5
) & theme(axis.title.x = element_blank())

print(pre_filter_violin)
```

#### MAD-based adaptive thresholding

```{r}
# Cells deviating >3 MADs from the median are flagged as outliers.
# Convert to SCE to use scuttle's QC infrastructure
sce <- as.SingleCellExperiment(sc_data)
# Compute per-cell QC metrics with mitochondrial and ribosomal subsets
cell_qc <- perCellQCMetrics(
  sce,
  subsets = list(
    Mito = grep("^MT-",    rownames(sce), value = FALSE),
    Ribo = grep("^RP[SL]", rownames(sce), value = FALSE)
  )
)

adaptive_filters <- quickPerCellQC(
  cell_qc,
  percent_subsets = "subsets_Mito_percent",
  nmads           = 3
)

# Inspect how many cells each filter removes independently
print(colSums(as.matrix(adaptive_filters)))
```

#### Apply a hard floor as a safety net

```{r}
hard_filter <- with(sc_data@meta.data,
  nFeature_RNA    > 300        &
  nFeature_RNA    < 5000        &
  nCount_RNA      > 500        &   
  nCount_RNA      < 30000        & 
  percent.mt      < 30         &    
  log10GenesPerUMI > 0.80           
)

cells_pass <- !adaptive_filters$discard & hard_filter

cat("Cells before filtering:", ncol(sc_data), "\n")
cat("Cells passing QC:       ", sum(cells_pass), "\n")
cat("Cells removed:          ", sum(!cells_pass), "\n")

sc_data_filtered <- sc_data[, cells_pass]
```

#### gene-level filtering on the clean cell set

```{r}
gene_cell_counts <- rowSums(GetAssayData(sc_data_filtered, layer = "counts") > 0)
genes_keep <- gene_cell_counts >= 10

cat("\nGenes before gene filtering:", nrow(sc_data_filtered), "\n")
cat("Genes retained (>=10 cells):", sum(genes_keep), "\n")

sc_data_clean <- sc_data_filtered[genes_keep, ]
```

#### Post-filtering QC check

```{r}
post_filter_violin <- VlnPlot(
  sc_data_clean,
  features = qc_features,
  pt.size  = 0,
  ncol     = 5
) & theme(axis.title.x = element_blank(),
          plot.title = element_text(size=10),
          axis.text.x = element_text(size=8))

print(post_filter_violin)

# Print a clean summary
cat("\n──── Final object summary ────\n")
cat("Cells:   ", ncol(sc_data_clean), "\n")
cat("Genes:   ", nrow(sc_data_clean), "\n")
cat("Metadata:", paste(colnames(sc_data_clean@meta.data), collapse = ", "), "\n")
```

#### Normalize data

```{r}
sc_normalize <- NormalizeData(sc_data_clean, 
                             normalization.method = "LogNormalize")

sc_normalize <- FindVariableFeatures(sc_normalize, 
                                    selection.method = "vst", 
                                    nfeatures = 2000)

top10 <- head(VariableFeatures(sc_normalize), 10)
plot1 <- VariableFeaturePlot(sc_normalize)
plot2 <- LabelPoints(plot = plot1, 
                     points = top10, 
                     repel = TRUE, 
                     xnudge = 0, 
                     ynudge = 0) +
  theme(legend.position = "top") +
  labs(title = "Top 2000 Variable Features")
plot2
```

#### dimension reduction
```{r}
sc_normalize <- ScaleData(sc_normalize, 
                         features = VariableFeatures(sc_normalize))
sc_normalize <- RunPCA(sc_normalize, 
                      features = VariableFeatures(sc_normalize), verbose = FALSE)
viz_pca <- VizDimLoadings(sc_normalize, dims = 1:2, reduction = "pca")
pca_plot <- DimPlot(sc_normalize, reduction = "pca", group.by = "orig.ident") + 
  labs(title = "PCA Reduction")
elbow_plot <- ElbowPlot(sc_normalize, ndims = 50) + 
  labs(title = "Elbow Plot: Selection of PCs")
pca_plot|elbow_plot
```

```{r}
pc_num <- 15
sc_normalize <- FindNeighbors(sc_normalize, dims = 1:pc_num)
sc_normalize <- FindClusters(sc_normalize, resolution = 0.4)
sc_normalize <- RunUMAP(sc_normalize, dims = 1:pc_num)
DimPlot(sc_normalize, reduction = "umap", label = TRUE)
```

#### cell type annotation based on scType program which have well-annotated HCC information
```{r}
# source("https://raw.githubusercontent.com/IanevskiAleksandr/sc-type/master/R/sctype_wrapper.R")
# sample <- run_sctype(sc_normalize, assay = "RNA", scaled = TRUE, known_tissue_type="Liver",custom_marker_file="https://raw.githubusercontent.com/IanevskiAleksandr/sc-type/master/ScTypeDB_short.xlsx",name="sctype_classification")
# sample <- subset(sample,sctype_classification != "Unknown")
sample <- readRDS(file = "./filtered_seuobj.rds")
DimPlot(sample, reduction = "umap", 
        group.by = "sctype_classification",
        label = TRUE,label.size = 3)
```

```{r,fig.width=7, fig.height=9}
Idents(sample) <- "sctype_classification"
# sc_markers <- FindAllMarkers(sample, 
#                              only.pos = TRUE, 
#                              min.pct = 0.1, 
#                              logfc.threshold = 0.25)

#saveRDS(sc_markers,file="cellType_markers.rds")
sc_markers <- readRDS("cellType_markers.rds")
top3_markers <- sc_markers %>%
  group_by(cluster) %>%
  top_n(n = 3, wt = avg_log2FC)

DotPlot(sample,features = top3_markers$gene,group.by = "sctype_classification")+
  theme(axis.text.x = element_text(angle=90,hjust=1))+
  coord_flip()+
  theme(axis.text=element_text(size=9))
  
```
#### process bulk RNA-seq data
```{r}
## to convert ENSEMBL ID to gene symbol
TCGA_FPKM<- read.table("TCGA-LIHC.star_fpkm.tsv.gz",
                       stringsAsFactors = T,row.names = 1,header=T)
colnames(TCGA_FPKM) <- str_replace_all(colnames(TCGA_FPKM),"\\.","-")

ensembl_ids <- rownames(TCGA_FPKM)
clean_ids <- gsub("\\..*", "", ensembl_ids) 
id_map <- bitr(clean_ids, 
               fromType = "ENSEMBL", 
               toType = "SYMBOL", 
               OrgDb = org.Hs.eg.db)

TCGA_df <- as.data.frame(TCGA_FPKM)
TCGA_df$ENSEMBL <- clean_ids
TCGA_final <- TCGA_df %>%
  inner_join(id_map,by="ENSEMBL") %>%
  dplyr::select(-ENSEMBL) %>%
  group_by(SYMBOL) %>%
  summarise(across(everything(), mean)) %>% 
  ungroup()
TCGA_final <- as.data.frame(TCGA_final)
rownames(TCGA_final) <- TCGA_final$SYMBOL
TCGA_final <- TCGA_final[, -1] 
saveRDS(TCGA_final,"TCGA_fpkm_symbol.rds")

raw_clinic_info <- read.table("TCGA-LIHC.clinical.tsv.gz",
                              header=T,sep="\t",quote="\"")
coldata <- data.frame(row.names=raw_clinic_info$sample,
                     condition=raw_clinic_info$sample_type.samples) %>%
  mutate(condition=factor(condition,
                          levels=c("Primary Tumor","Solid Tissue Normal"),
                          labels=c("Tumor","Control")))
```

#### cell type deconvolution based on BisqueRNA

```{r}
# library(BisqueRNA)
# bulk.eset <- Biobase::ExpressionSet(assayData = as.matrix(TCGA_final))
# sc.eset <- BisqueRNA::SeuratToExpressionSet(sample, delimiter="_", position=1, version="v3")
# res <- BisqueRNA::ReferenceBasedDecomposition(bulk.eset, sc.eset, markers=NULL, use.overlap=FALSE)
# ref.based.estimates <- res$bulk.props
# saveRDS(ref.based.estimates,file="ref.based.estimates.rds")
ref.based.estimates <- readRDS("ref.based.estimates.rds")
ref.based.estimates[1:6,1:6]
```

#### implement comparison of cell proportion between tumor and control samples
```{r}
long_data <- ref.based.estimates %>%
  t() %>%
  as.data.frame() %>%
  mutate(sample=row.names(.)) %>%
  pivot_longer(cols=1:15) %>%
  mutate(group=coldata[sample,]) %>%
  filter(!is.na(group))

long_data %>%
  mutate(group=factor(group,levels=c("Control","Tumor"))) %>%
  ggplot(aes(x=group,y=value,color=group))+
  geom_boxplot(fill="NA")+
  facet_wrap(.~name,ncol = 5,scales = "free")+
  stat_compare_means(label = "p.signif",label.y.npc = 0.85)+
  scale_color_manual(values = c("steelblue","orange"))+
  theme_bw()+
  theme(legend.position = "NA",
        strip.text = element_text(size=7),
        strip.background = element_blank())
  
```
This result shows that all cell types are differently proportioned between tumor and control samples.

#### cell proportion across all samples exhibition
```{r}
my_colors <- colorRampPalette(brewer.pal(12, "Paired"))(22)
long_data <- long_data %>% arrange(group,name) %>%
  mutate(group=factor(group,levels=c("Control","Tumor")),sample=factor(sample,levels=unique(sample)))
p_anno <- ggplot(long_data, aes(x = sample, y = 1, fill = group)) +
  geom_tile() + 
  scale_fill_manual(values=c("#4DBBD5FF","#E64B35FF")) +
  theme_void() + 
  theme(legend.position = "top") +
  labs(fill = "Group")

p_bar <- ggplot(long_data, aes(x = sample, y = value, fill = name)) +
  geom_bar(stat = "identity", width = 1) +
  scale_fill_manual(values = my_colors) +
  scale_y_continuous(expand = c(0, 0)) +
  theme_classic() +
  theme(
    axis.text.x = element_blank(), 
    axis.ticks.x = element_blank(),
    legend.position = "bottom"
  ) +
  labs(y = "Cell Proportion", fill = "Cell Type")

p <- p_anno / p_bar + plot_layout(heights = c(1, 10))

p

```
The barplot shows that the proportion of cells are different between control and tumor samples, for instance, the macrophages take higher proportion in control samples.

#### cell type proportion correlation 
```{r}
cor_res <- ref.based.estimates %>%
  t() %>%
  as.data.frame() %>%
  corr.test(method = "pearson")
cor_matrix <- cor_res$r     
p_matrix <- cor_res$p  
corrplot(cor_matrix, 
         method = "color", 
         type = "upper", 
         order = "hclust",  
         p.mat = p_matrix, 
         sig.level = 0.05, 
         insig = "label_sig", 
         tl.col = "black", 
         tl.srt = 45,
         tl.cex = 0.7,
         cl.cex = 0.7,
         pch.cex = 0.8,
         col = colorRampPalette(c("#4477AA", "#77AADD", "#FFFFFF", "#EE9988", "#BB4444"))(200))

```

#### exploring the cell proportion change along stage evolution
```{r}
stage_anno <- raw_clinic_info %>%
  dplyr::select(sample,ajcc_pathologic_stage.diagnoses) %>%
  mutate(ajcc_pathologic_stage.diagnoses= case_when(ajcc_pathologic_stage.diagnoses %in% c("Stage IIIA","Stage IIIB","Stage IIIC") ~ "Stage III",
            ajcc_pathologic_stage.diagnoses %in% c("Stage IVA","Stage IVB") ~ "Stage IV",
            .default = ajcc_pathologic_stage.diagnoses)) %>%
  filter(ajcc_pathologic_stage.diagnoses != "") %>% 
  column_to_rownames(var="sample")

plot_data2 <- long_data
plot_data2$stage <-   stage_anno[plot_data2$sample,]
plot_data2$stage <- factor(plot_data2$stage,levels = c("Stage I","Stage II","Stage III","Stage IV"))
plot_data2 <- plot_data2 %>% filter(group == "Tumor",!is.na(stage), stage != "")

plot_data2 <- plot_data2 %>%
  mutate(stage_num = as.numeric(stage)) # Stage I -> 1, Stage II -> 2 ...

my_colors <- colorRampPalette(brewer.pal(12, "Paired"))(15)

p <- ggplot(plot_data2, aes(x = stage_num, y = value, color = name)) +
  geom_boxplot(aes(group = interaction(stage, name), fill = name), color = "black") + 
  geom_smooth(method = "gam", formula = y ~ s(x, k = 3), se = FALSE, size = 1) + 
  scale_color_manual(values=my_colors) + 
  scale_fill_manual(values=my_colors) + 
  scale_x_continuous(breaks = 1:4, labels = c("Stage I", "Stage II", "Stage III", "Stage IV")) +
  labs(x="",y="cell proportion")+
  theme_bw()+
  theme(
    axis.text = element_text(size = 8, color="black"),
    plot.title = element_text(size = 10, hjust = 0.5, face = "bold"),
    legend.text = element_text(size = 7),
    legend.position = "right"
  )
p
```
There are certain patterns of cell proportion change along tumor progression, for instance, macrophages tends to take less proportion in the early stage of tumor progression in stage II and III, but starting to rise in stage IV. However, platelets decreases along the progression stages.

#### process the raw counts of bulk RNA-seq data and perform DEG analysis
```{r fig.width= 5,fig.height=5}
raw_counts <- read.table("TCGA-LIHC.star_counts.tsv.gz",
                         stringsAsFactors = F,header=T,row.names=1)
colnames(raw_counts) <- str_replace_all(colnames(raw_counts),pattern = "\\.",replacement = "-")
coldata <- coldata %>% filter(!is.na(condition))
common_sample <- intersect(colnames(raw_counts),row.names(coldata))
raw_counts <- raw_counts[,common_sample]
coldata_new <- coldata[common_sample,]
coldata_new <- data.frame(row.names = common_sample,condition=coldata_new)
clean_ids <- gsub("\\..*", "", rownames(raw_counts))
gene_map <- bitr(clean_ids, 
                 fromType = "ENSEMBL", 
                 toType = "SYMBOL", 
                 OrgDb = org.Hs.eg.db)


keep_indices <- which(clean_ids %in% gene_map$ENSEMBL)
raw_counts_filtered <- raw_counts[keep_indices, ]

final_clean_ids <- gsub("\\..*", "", rownames(raw_counts_filtered))
final_map <- gene_map[match(final_clean_ids, gene_map$ENSEMBL), ]

final_symbols <- make.unique(final_map$SYMBOL)
rownames(raw_counts_filtered) <- final_symbols
raw_counts_filtered <- raw_counts_filtered^2 -1
raw_counts_filtered[raw_counts_filtered<0] <- 0
write.csv(raw_counts_filtered,"raw_count_symbol.csv")

dds <- DESeqDataSetFromMatrix(countData = round(raw_counts_filtered,0),
                              colData = coldata_new,
                              design = ~ condition)
dds <- DESeq(dds)

saveRDS(dds,"dds.rds")
res <- results(dds, contrast = c("condition", "Tumor", "Control"))
FDR_THRESHOLD <- 0.05
LFC_THRESHOLD <- 1

# Volcano Plot
generate_volcano_plot <- function(results_df, title) { 
  plot_df <- results_df %>% 
    filter(!is.na(pvalue)) %>%
    rownames_to_column(var = "Gene_Symbol") %>% 
    mutate(
      Significant = case_when(pvalue < FDR_THRESHOLD & log2FoldChange >= LFC_THRESHOLD ~ "Up-regulated", 
                              pvalue < FDR_THRESHOLD & log2FoldChange <= -LFC_THRESHOLD ~ "Down-regulated", 
                              TRUE ~ "Not significant"),
      log10_pvalue = -log10(pvalue+0.00000001)
    )
  # 找到数据中最大的有限数值
  max_finite <- max(plot_df$log10_pvalue[is.finite(plot_df$log10_pvalue)], na.rm = TRUE)
  
  plot_df <- plot_df %>%
    mutate(log10_pvalue = ifelse(is.infinite(log10_pvalue), max_finite * 1.05, log10_pvalue))
  
  top10_up <- plot_df %>% 
    filter(Significant == "Up-regulated") %>% 
    arrange(desc(log2FoldChange)) %>% 
    head(10)
  top10_down <- plot_df %>% 
    filter(Significant == "Down-regulated") %>% 
    arrange(log2FoldChange) %>% head(10)
  
  top_genes_label <- bind_rows(top10_up, top10_down)
  
  p <- ggplot(plot_df, aes(x = log2FoldChange, y = log10_pvalue, color = Significant)) +
    geom_point(alpha = 0.5, size = 1.5) +
    scale_color_manual(values = c("Up-regulated" = "red", "Down-regulated" = "blue", "Not significant" = "grey50")) +
    geom_hline(yintercept = -log10(FDR_THRESHOLD), linetype = "dashed", color = "black", alpha = 0.5) +
    geom_vline(xintercept = c(-LFC_THRESHOLD, LFC_THRESHOLD), linetype = "dashed", color = "black", alpha = 0.5) +
    geom_text_repel(data = top_genes_label, aes(label = Gene_Symbol), 
                    size = 3, box.padding = unit(0.5, "lines"), 
                    point.padding = unit(0.5, "lines"), max.overlaps = 30) +
    labs(title = paste0( "Volcano Plot: ", title), x = expression(Log[2]~"Fold Change"), y = expression(-Log[10]~"P-value")) +
    scale_y_continuous(limits = c(0,1.2 * max(plot_df$log10_pvalue)))+
    theme_minimal(base_size = 14) + 
    theme(legend.position = "bottom",
          plot.title = element_text(hjust = 0.5, face = "bold"))
  
  return(p)
}
generate_volcano_plot(as.data.frame(res), "Tumor vs. Normal" )
```
#### gene enrichment analysis
```{r}
# Create a named list of cell-type gene sets from sc markers
sc_markers_new <- sc_markers %>%
  filter(p_val_adj<0.05,avg_log2FC>0.5)
cell_type_genesets <- split(sc_markers_new$gene, sc_markers_new$cluster)

bulk_ranked <- res$log2FoldChange
names(bulk_ranked) <- rownames(res)
bulk_ranked <- sort(bulk_ranked, decreasing = TRUE)
# Run GSEA
gsea_result <- GSEA(
  geneList     = bulk_ranked,
  TERM2GENE    = stack(cell_type_genesets)[, c("ind", "values")],
  pvalueCutoff = 0.05,
  verbose      = FALSE
)
```
